---
title: "Preparing the data"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Preparing the data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(vizsurvey)
```

# Introduction

This vignette presents the data preparation required to use {vizsurvey} through the `runVizsurvey_from_folder` call. The goal is to enable regular, structured use of the tool within a team managing several surveys.

In this context, launching the interface manually for each database quickly becomes tedious. Moreover, computing the gaps, especially when they are broken down by survey wave or by filter, can be resource-intensive.

To make this process easier, {vizsurvey} offers several automated preparation functions. This vignette describes how they work, step by step, from structuring the files to generating the objects needed to launch the application.

# Example data

EU-SILC is an annual survey coordinated by Eurostat. It is used to measure income, poverty, inequality and living conditions in European countries. Eurostat provides [Public Use Files](https://ec.europa.eu/eurostat/web/microdata/public-microdata) that can be used freely for testing. The files are anonymised and do not necessarily reproduce the official results. We provide two years of the EU-SILC survey for Belgium, for the household file. There is no interviewer number in these files initially, so a fictitious number `NR_ITW` was created for the examples here.

The files can be found in the package source files at `inst/extdata/SILC`, or downloaded [here](https://github.com/tdelc/vizsurvey/tree/master/inst/extdata/SILC/).

# Configuring a survey

Your database file must be readable by the `data.table::fread` function. This function automatically recognises standard formats (CSV, TSV, etc.). Place your file in a dedicated directory, where it will be the **only file of that format**.

If your survey is made up of several files (for example one file per year or per country, as for the EU-SILC survey), you can place them in the same directory. All files must, however, contain the same variables as those listed in the configuration file, and be in the same format. In addition, no other file of that type must be present in the directory.

## Survey configuration file

Each data directory must contain a configuration file named `config.txt`. This file tells {vizsurvey} how to interpret and structure your data. It must include the following elements:

```         
name_survey = 

vars_discretes =
vars_continous =

var_wave = 
var_filter = 

var_intvwr = 
```

-   `name_survey`: name of the survey (useful for identifying the files);

-   `vars_discretes`: variables to be treated as categorical;

-   `vars_continous`: variables to be treated as continuous;

-   `var_wave`: survey wave variable used to break down the calculations (example: year);

-   `var_filter`: filter variable allowing an additional breakdown (example: province). The calculations are also produced without any filter distinction;

-   `var_intvwr`: interviewer identifier variable.

The file can be created manually, or generated automatically with the `create_config` function. Here is the example for the household files of the SILC survey, with the year (`HB010`) as the survey wave and the province (`HB020`) as the filter:

```{r, eval=F, warning = F, message = F}
create_config(
  folder_path = "inst/extdata/SILC/HFILE",
  name_survey = "SILC-H",
  var_wave    = "HB010",
  var_filter    = "HB020",
  var_intvwr   = "NR_ITW"
)
```

The discrete and continuous variables only need to be specified if the `classify_df` function misidentifies their type. This function automatically determines the type of each variable, based on its format and number of categories. The threshold for classifying a variable as categorical is set to 15 by default.

```{r}
classify_df(iris)
```

## Optional: interviewer summary file

In addition to the microdata, you can place an **optional summary file containing one row per interviewer** in the survey directory. This file gathers interviewer-level indicators that are not derived from the response distributions themselves, for example the number of completed interviews, the average interview duration, or the response rate. It must include the interviewer identifier (the same one used in `var_intvwr`) so that it can be matched to the microdata.

To use it, declare its filename in `config.txt` with the `file_synthesis` key, or pass it to `create_config`:

```         
file_synthesis = synthesis.csv
```

```{r, eval=F, warning = F, message = F}
create_config(
  folder_path    = "inst/extdata/SILC/HFILE",
  name_survey    = "SILC-H",
  var_wave       = "HB010",
  var_filter     = "HB020",
  var_intvwr     = "NR_ITW",
  file_synthesis = "synthesis.csv"
)
```

When such a file is present, it is automatically integrated and analysed during preparation, so that any atypical interviewer-level indicator is flagged as a potential anomaly, in the same way as the distributional gaps.

# Preparing the survey

## Simple structure: a single directory

Once the configuration is in place, {vizsurvey} can compute **in advance** all the statistics needed to monitor the survey. When you have only a single survey directory, the `prepa_survey` function uses this directory, containing the data and the `config.txt` file, to produce:

-   per wave, the descriptive statistics of each variable;

-   per wave, for each variable, the gaps of each interviewer relative to the population (within the wave);

-   per wave and per filter, for each variable, the gaps of each interviewer relative to the population (within the filter and the wave).

By default, `prepa_survey` looks for the CSV files in the directory, but the `file_pattern` argument lets you adapt this behaviour. A `global.rds` file is then generated in the same folder: it is the only file that will be used by the interactive interface.

```{r, eval=F, warning = F, message = F}
prepa_survey(
  folder_path  = "inst/extdata/SILC/HFILE",
  file_pattern = "*.csv")
```

The application can then be launched simply with the `runVizsurvey_from_folder` function.

```{r, eval=F, warning = F, message = F}
runVizsurvey_from_folder("inst/extdata/SILC/HFILE", depth_folder = 1)
```

The `prepa_survey` function can be called each time your survey data are updated. Its execution can be scheduled so that all the statistics on your surveys are computed daily.

## Double structure: several directories

If you manage several surveys, you can create the directories side by side following this structure:

```         
data/
  ├── ENQ1/
  │   ├── *.csv
  │   └── config.txt
  └── ENQ2/
      ├── *.csv
      └── config.txt
```

You can then prepare all your surveys with the `prepa_surveys` function. This function acts as a wrapper around `prepa_survey` and automatically applies the preparation to all the child directories of the specified path. In this case, you need to set `depth_folder = 2`.

```{r, eval=F, warning = F, message = F}
prepa_surveys(folder_path  = "inst/extdata/SILC", depth_folder = 2)
```

You can then launch the interface by specifying the same depth level:

```{r, eval=F, warning = F, message = F}
runVizsurvey_from_folder("inst/extdata/SILC", depth_folder = 2)
```

## Triple structure: several directory levels

It is also possible to manage a full survey tree, with several hierarchical levels. The first level may correspond to the different surveys conducted and, as a second level, one directory per type of file produced by that survey. For example, the SILC survey contains four types of files, two at the household level and two at the individual level. The HBS survey contains a survey file and a file for expenditure diaries, and so on. Here is an example of the expected directory structure:

```         
data/
  ├── ENQ1/
  │   ├── ENQ1-A/
  │   │   └── ...
  │   └── ENQ1-B/
  │       └── ...
  └── ENQ2/
      ├── ENQ2-A/
      │   └── ...
      └── ENQ2-B/
          └── ...
```

Each sub-directory must contain its own `config.txt` and its own data files. Within each sub-directory, all files must have the same data structure. You can then run the global preparation:

```{r, eval=F, warning = F, message = F}
prepa_surveys(folder_path  = "data", depth_folder = 3)
```

Finally, the interface is launched with the same depth:

```{r, eval=F, warning = F, message = F}
runVizsurvey_from_folder("data", depth_folder = 3)
```

If the data are updated, for example during fieldwork, you simply need to run the `prepa_surveys` function again to obtain an updated version of the `global.rds` objects, and thus a full update of the interactive interface.

# Conclusion

Data preparation is an essential step before fully using {vizsurvey} within an organisation that runs surveys on a daily basis. It optimises the responsiveness of the application and guarantees that all survey managers analyse the same up-to-date files.

Once the files are structured, configured and prepared, each launch of the interface becomes immediate: all the necessary statistics are already computed and stored in the `global.rds` file. This approach ensures an efficient reuse of the data, whatever the number of surveys or the complexity of their tree structure. It also facilitates collaborative work within a team: each member can explore the results without having to know R.
