---
title: "Methodology and use of the functions"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Methodology and use of the functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = F, message = F}
library(vizsurvey)
library(isotree)
library(tibble)
library(dplyr)
library(tidyr)
library(plotly)
```

# Introduction

This vignette presents the methodological foundations and the calculations performed by the {vizsurvey} package. As explained in the first vignette, {vizsurvey} is not meant to run statistical tests or to draw conclusions: it is above all an exploration and diagnostic tool, positioned between data collection and analysis in the GSBPM cycle.

The calculations carried out in {vizsurvey} rely on simple but robust measures, such as distances, ranks, standard deviations or relative gaps between distributions. These measures serve as visual cues to identify possible inconsistencies: abnormal variations from one wave to the next, gaps between interviewers, abrupt changes in distribution, or unexpected increases in the non-response rate.

All the functions used in {vizsurvey} were designed to be applicable to any type of variable, whether categorical (codes, categories, multiple-choice answers) or numeric (continuous values, durations, ages, incomes). The aim is to provide a generic, reproducible and robust computation framework that ensures the comparability of results across surveys and survey waves. Priority is given to the stability of the indicators and to the simplicity of the formulas, so as to make them easy to read and to adopt for field teams.

These methodological choices are not set in stone: they were progressively refined within Statbel, based on the first results produced by the interface. Feedback from survey managers, exchanges between departments, and several references from the survey-quality literature have guided the evolution of the functions.

Publishing this package is a way to share these choices. By making the formulas, functions and computation logic public, we aim to promote transparency, encourage reuse and stimulate contributions from other teams facing similar issues. Our ambition is to build, together with other statistical institutes and polling organisations, a common base of tools and practices for survey monitoring.

The examples provided in this vignette are based on the `eusilc` dataset from the {laeken} package. This file does not contain an interviewer identifier, so we create one from the province variable. This lets us simulate the presence of 45 interviewers.

```{r}
library(laeken)
data(eusilc)
set.seed(123)
eusilc$NR_ITW <- paste(eusilc$db040, sample(1:5, nrow(eusilc), replace = T), sep = "-")

table(eusilc$NR_ITW)
```

# `classify_df()`: classify the variables

`classify_df()` takes a database as input and produces a table as output, with a category for each variable: "Solo" if the variable contains only one category; "Modal" when the variable contains at most 15 categories; "Continuous" when the variable is numeric with more than 15 categories; "Text" otherwise. The number of categories used to choose the variable's category can be set with the `threshold` parameter (15 by default).

This function guides the {vizsurvey} analysis in choosing the indicators for each variable. You can let the function decide the category of your variables on its own, or force the category of certain variables yourself with `config.txt` (see vignette 2).

```{r}
classify_df(eusilc)
```

# `prepa_stats()`: extract the aggregates

`prepa_stats()` takes as input a database and an interviewer variable, and returns a data.frame compiling the statistics computed for each variable, per interviewer. You can indicate the category of each variable with the arguments `vars_vd` (discrete variables) and `vars_vc` (continuous variables). If both arguments are left empty, `classify_df` is run on the database before computing the statistics.

```{r eval=FALSE}
prepa_stats(eusilc, var_intvwr = "NR_ITW")
```

## Description of the output

This function returns a long file, with one row per interviewer/variable/indicator crossing. For each interviewer/variable crossing, the number of rows, the number of valid (non-missing) rows and the variable type are reported. For a given interviewer, the indicators computed are:

-   "missing": percentage of missing values in the variable;
-   "presence": an indicator of whether at least one value is non-missing;
-   "Nmod" (for categorical variables): number of categories of the variable;
-   "chi2" (for categorical variables): distance measure of the variable for this interviewer relative to the full database;
-   "mean" / "median" (for continuous variables): mean / median of the variable.

The `value` variable gives the value of the indicator for the interviewer. The `value_ref` variable gives the value of the indicator for the whole population. The `standard` variable is a standardisation of the variable (detailed below).

## Computing the $\chi^2$

Most indicators have simple formulas. Only the "chi2" indicator deserves particular attention. Here are the steps for computing the $\chi^2$:

1.  We start by computing the distribution of the variable's categories over all rows, using the `list_dist` function. We build a table of counts per category, then convert it into proportions. We include an "NA" category when the variable contains missing data. Rare categories (less than 1%) are grouped into an "Other" category. We obtain an `expected_prop` object for each variable;

2.  Within each interviewer, the `my_chisq_test` function identifies the missing values and the rare categories (listed in the previous step) in order to group them;

3.  We compare the interviewer's distribution with that of the full population, using `stats::chisq.test(obs, p = exp_prop)`.

Grouping categories below 1% avoids visual noise and irrelevant comparisons. Treating missing data as a dedicated category makes it possible to track non-responses on the same footing as valid responses.

Here, an interviewer is partly compared with themselves, because the overall distribution is computed only once. This can be a problem if the interviewer has too much influence on the full database. We made this choice to limit computation time. Only one distribution has to be computed per variable, and {vizsurvey} then makes it possible to break these calculations down by wave and by filter.

## Standardising the indicators

The standardisation of the indicators is performed with the `scale_IQR` function. The function aims to bring numeric variables onto a comparable scale, but replaces the usual standard deviation with the interquartile range (IQR). It is therefore a robust normalisation, less sensitive to extreme values than the usual standardisation. It is designed to fall back on the classic `scale` standardisation in exceptional situations: missing values, lack of variability, or data bounded between 0 and 1.

# `heatmap_group()`: customise the visualisation

You can generate the heatmap outside the interactive {vizsurvey} interface. The `heatmap_group` function requires a data.frame produced by `prepa_stats` and returns a ggplot with one row per interviewer and one column per variable. The function uses all the indicators available for a given variable to determine the colour intensity of the cell. If you want to limit the analysis to certain indicators, filter the output of `prepa_stats` before computing the heatmap.

By default, cells are displayed when the standardised gap exceeds 5; you can change this threshold with the `threshold` argument. The colour goes from white to the colour selected by the `color` argument (red by default), on a scale from the threshold to twice that value. This way, an excessively large gap does not hide the other gaps.

```{r}
df_stats_eusilc <- prepa_stats(eusilc, "NR_ITW")

heatmap_group(df_stats_eusilc, threshold = 1, color = "green2") +
  ggtitle("eusilc heatmap")
df_stats_eusilc %>% 
  filter(stat == 'chi2') %>% 
  heatmap_group(threshold = 1, color = "purple2") +
  ggtitle("Heatmap of eusilc chi²")
```

You can also zoom into the heatmap, again by filtering the data.frame produced by `prepa_stats`. Here are the three examples used in the interactive interface:

```{r}
threshold <- 3

df_stats_eusilc %>%
  group_by(NR_ITW) %>%
  filter(max(abs(standard), na.rm = TRUE) > threshold) %>%
  ungroup() %>% 
  heatmap_group(threshold) +
  ggtitle("Zoom on rows")
df_stats_eusilc %>%
  group_by(variable) %>%
  filter(max(abs(standard), na.rm = TRUE) > threshold) %>%
  ungroup() %>% 
  heatmap_group(threshold) +
  ggtitle("Zoom on columns")
df_stats_eusilc %>%
  group_by(NR_ITW, variable) %>%
  filter(max(abs(standard), na.rm = TRUE) > threshold) %>%
  ungroup() %>% 
  heatmap_group(threshold) +
  ggtitle("Zoom on cells")
```

The ggplot object produced by `heatmap_group` embeds tooltips that can be activated with {plotly}.

```{r}
p <- df_stats_eusilc %>%
  heatmap_group(3) +
  ggtitle("Interactive heatmap")

ggplotly(p, tooltip = "text")
```

# `score_isoforest()`: rank the atypical profiles

The `score_isoforest` function aims to detect atypical observations in a multivariate dataset, relying on the [Isolation Forest](https://cran.r-project.org/web/packages/isotree/vignettes/An_Introduction_to_Isolation_Forests.html) method. This is an unsupervised learning algorithm, particularly useful for spotting individuals or records whose profile differs sharply from the rest of the population. This function is designed to be applied to a set of numeric variables describing the same type of statistical unit (here, interviewers), and returns an isolation score for each observation.

We propose applying the method to the indicators computed for each variable by the `prepa_stats` function. For now, we keep the standardised values of the $\chi^2$ for categorical variables and of the medians for continuous variables. The `score_isoforest` function requires a wide file, so we pivot the indicators before applying it.

```{r}
df_prepa <- df_stats_eusilc %>%
  filter(stat %in% c("chi2", "median")) %>%
  select(NR_ITW, variable, standard) %>%
  pivot_wider(
    id_cols = NR_ITW,
    names_from = variable,
    values_from = standard
  ) %>% 
  column_to_rownames(var = "NR_ITW")

head(sort(score_isoforest(df_prepa), decreasing = T))
hist(score_isoforest(df_prepa))
```

If an optional interviewer summary file has been supplied (one row per interviewer, see the first vignette), its standardised indicators are appended as additional columns to this wide table, so that interviewer-level anomalies (for example an unusually short average interview duration or an extreme response rate) are scored together with the distributional ones.

We again find the interviewers from the province of Vienna coming out on top. A threshold of 0.5 seems acceptable to decide that a score reveals an anomaly, according to [this article](https://scispace.com/pdf/an-in-depth-study-and-improvement-of-isolation-forest-204k9g39.pdf). Once again, {vizsurvey} does not aim to propose a single threshold, but rather a ranking to guide the investigation.

# Conclusion

All the functions presented in this vignette rest on a simple idea: make the weak signals in survey data visible, without over-interpreting them. By combining robust indicators (interquartile ranges, $\chi^2$, isolation scores) and intuitive visual representations (heatmaps, rankings), the tool makes it possible to quickly spot unexpected gaps, whether they are due to data-entry problems, interpretation errors, context effects or real phenomena.

The goal of {vizsurvey} is also collaborative. By documenting the formulas and making the methods transparent, we want to encourage discussion around the monitoring of survey data, share good practices and pave the way for collective improvements. The methodological choices presented here are not a definitive standard, and will evolve based on feedback on the package.
